The Gravettian-Solutrean transition in westernmost Iberia: new data from the sites of Vale Boi and Lapa do Picareiro

### 2020-09-21 analysis/paper/paper.Rmd
- Updated the file to correspond to article corrections

---
title: "The Gravettian-Solutrean transition in westernmost Iberia: new data from the sites of Vale Boi and Lapa do Picareiro"
author:
  - Joana Belmiro: 
      email: jnetobelmiro@gmail.com
      institute: [ICArEHB]
      correspondence: true
  - João Cascalheira:
      email: jmcascalheira@ualg.pt
      institute: [ICArEHB]
      correspondence: false
  - Nuno Bicho:
      email:
      institute: [ICArEHB]
  - Jonathan Haws:
      email:
      institute: [UL, ICArEHB]
institute:
  - ICArEHB: ICArEHB, Universidade do Algarve
  - UL: University of Louisville
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: bibliography.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  This study presents the analysis of the lithic assemblages from Layers 5 and 4E (Terrace) of the site of Vale Boi (southern Portugal) and Levels U and T from Lapa do Picareiro (central Portugal). We aimed to understand the technological patterns and raw material exploitation during the Gravettian-Solutrean transition in westernmost Iberia, and test the traditional models with assemblages from recently excavated sites, while expanding the geographic range.
  Results show the existence of two discrete phases in each site. The first, with high frequency of quartz use for bladelet production, seems to reflect the presence, in both sites, of a Terminal Gravettian horizon, as defined by Almeida (2000). The second, with some significant differences between sites, attests the presence of Vale Comprido technology and lower quartz frequencies at Vale Boi, representing a Proto-Solutrean occupation; and the presence of a blade component in Lapa do Picareiro that, together with the respective absolute chronology, may be attributed to a Proto-Solutrean or an Early Solutrean horizon.
  In general terms, this study allowed to confirm that the Terminal Gravettian and the Proto-Solutrean are discrete phases across the transiton, in concordance with the Three-Phase model presented by Zilhão [-@zilhao1997]. It further consolidates the expansion of similar techno-cultural patterns to southern Portugal, that may be explained by the growth of social networks [@cascalheiraandbicho2013].

@@

The transition from the Gravettian to the Solutrean, impacted by adverse climatic conditions during the Heinrich Event 2 (HE2) and the onset of the Last Glacial Maximum (LGM), continues to be a critical topic to understand Upper Paleolithic human adaptations to environmental changes in Westernmost Europe.

There has been a significant amount of proposals suggesting a full synchronism between the onset of each Upper Paleolithic technocomplex and the occurrence of the most severe climate events [e.g., @bradtmoller2012;@cascalheiraandbicho2013; @bichoetal2017]. Bradtmöller et al. [-@bradtmoller2012] suggest a direct relationship between three of the Heinrich Events (4, 3 and 2) and the substitution of Neanderthal populations with the emergence of the Aurignacian, the appearance of the Gravettian, and the development of the Solutrean, respectively. Based on the theoretical framework of Panarchy [@holling2002], the authors propose the Repeated Replacement Model (RRM), where the HEs are understood as the primary climatic triggers for population turnover, through the breakdown of communication networks and cultural traditions, which were subsequently reorganized under different socio-cultural conditions [@bradtmoller2012].

In this framework, the rapid climatic changes brought by the HE2 (c. 26.4-24.3 ka cal BP - [@sanchez-goni2010]) were the trigger for the reorganization of human groups at the end of the Gravettian. Records indicate that during HEs, and with particular severity in the HE2, western and southern Iberia must have experienced abrupt environmental changes, which altered the location, abundance and availability of resources for the communities of hunter-gatherers [@cascalheiraandbicho2013]. In some parts of Iberia and southwestern France these changes seem to be accompanied by significant techno-typological modifications, combining both elements from the previous and succeeding phases. For some authors, the temporal and spatial patterning of the elements that mark the transiton is, however, sufficiently significant to individualized different technocomplexes - such as the case of the Proto-Solutrean [@cascalheiraandbicho2013; @renard2011; @zilhaoetal1995; @zilhao1997]. Still, although the Proto-Solutrean is an essential piece to understand how hunter-gatherer communities adapted, reinvented and abandoned their social and technological organization, most of what is known for the this technocomplex is geographically constricted and lacking good chronological markers [@cascalheiraandbicho2013]. This in turn  hampers the understanding of the Proto-Solutrean technological solutions and their role in the Gravettian-Solutrean transition.

Result of the study of materials from old excavations by Manuel Heleno in Rio Maior (Portuguese Estremadura) in the 1980s, and new archaeological work in the same region, the Proto-Solutrean has been described as a transitional technocomplex from Gravettian to Solutrean technologies, through a process of local development, but synchronous across southwestern Europe [@zilhaoetal1999].

In the westermost edge of Iberia this technocomplex is best known from both open-air and cave sites located in the Portuguese Estremadura (Figure \@ref(fig:protomap)).

@@

Here, the Proto-Solutrean has been integrated into the Gravettian-Solutrean transition as part of a sequence of techno-typological modifications that could have occurred across two or three different stages [@zilhao1994; @zilhao1997; @zilhaoetal1999].

The Two-stage model starts with a Final Gravettian phase, characterized by a moderate use of quartz (~15%), production of truncated backed bladelets and proto-magdalenian retouched blades [@zilhao1997]. The second phase is the Proto-Solutrean, frequently characterized by a high percentage of quartz use (~30%), rare presence of backed bladelets, and the production of Vale Comprido points, the technocomplex’s index fossil [@almeida2000; @zilhao1997; @zilhaoetal1995]. For the production of these tools, the Proto-Solutrean shows three different operative sequences: one for the production of Vale Comprido points, through the removal of elongated blanks with convergent profiles and thick platforms; another for the production of blades, of Gravettian tradition; and another for the production of bladelets, probably obtained through the exploitation of thick endscrapers or carinated elements [@zilhao1997].

Vale Comprido points are described as robust pieces, often characterized by convergent shapes, triangular cross-sections and plain platforms, often having a high elongation ratio, although not necessarily falling into the blade category. The Vale Comprido points have been used to describe the technological transition between the Proto-Solutrean and the Solutrean. In this case, these points are seen as an element of discontinuity with the previous technocomplex, where the organic points armed with microliths from the Gravettian are replaced by lithic points, with enough similarities to the *pointes à face plane* of the Middle Solutrean to be understood as a technological development [@zilhao2013].

The Three-stage model maintains the Final Gravettian phase and its defining characteristics but subdivides the following Proto-Solutrean stage in two. As such, there is an intermediate stage characterized by the intensive use of quartz (~30%), which corresponds to Laugerie-Haute’s Aurignatian V, and named Terminal Gravettian. The third stage is the Proto-Solutrean, where quartz use diminishes. In this model, the Vale Comprido points and associated reduction sequence may appear in either of the last two stages [@zilhao1997]. Alternatively, in the Two-phase model, the Terminal Gravettian is understood as a functional facies of the Proto-Solutrean phase, related to specialized occupations and production activities within the technocomplex.

Regardless of which model is accepted, a high degree of lithic technological variability seems to be present across the transition [@almeida2000; @zilhao1997; @zilhaoetal1999]. This variability has been interpreted by Cascalheira and Bicho [-@cascalheiraandbicho2013] as a technological race in response to the environmental modifications taking course during the HE2. The authors suggest that, in order to correspond to the external pressures, there may have been the need to diversify the economic strategies in use until that moment. The high exploitation of quartz, for example, formerly a secondary raw material, and the use of similar reduction strategies between quartz and chert, might represent one such economic response. The same possibly applies to the use of unprecedented raw materials, as is the case of the Proto-Solutrean levels in Vale Boi, where a new raw material starts to be used mainly for the manufacture of Vale Comprido points [@belmiro2017; @marreiros2009]. Likewise, alterations recorded in territoriality patterns during the Proto-Solutrean can be interpreted has extensive regional networks and may be related to modifications in the landscape [@cascalheiraandbicho2013; @cascalheira2019].

However, despite the existence of a rather comprehensive set of literature about the technocomplex in Central Portugal, which has allowed for a better understanding of the impacts of HE2 in hunter-gatherer communities, further studies are needed to fill the existing gaps. One of these gaps is the concentration of data regarding this technocomplex in the Portuguese Estremadura, whereas the Proto-Solutrean is still relatively unknown in other places throughout southwestern Europe. In the case of Portugal, only one other site in the south (Figure \@ref(fig:protomap)) has so far revealed the presence of this techno-complex. Another issue is the small amount of absolute dating for the Gravettian-Solutrean transition, which limits the chronological definition of the technocomplexes, and thus the testing of which transition model (if any) applies best [@cascalheiraandbicho2013].

Addressing these issues, through the study of lithic assemblages from sites with good stratigraphic preservation, allowing accurate absolute dating and spatial tracking, will undoubtedly help further understand the Gravettian-Solutrean transition, its patterns, stages and possible regional variations.

@@

By combining techno-typological information with stratigraphy data and absolute chronology, we present new data on: 1) the technological organization of levels 5/4E of Vale Boi, and levels U/T from Lapa do Picareiro; and 2) a stratigraphical and chronological separation between Terminal Gravettian horizons and more recent occupations at both sites, revealing the application of the same general set of patterns to geographic areas outside of the Portuguese Estremadura.

@@

For the lithic analysis we adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analyzed followed those present in specialized literature, such as @brezillon1968 and @tixier1980, paired with other Upper Paleolithic lithic attribute analyses [see e.g. @bicho1992, @zilhao1997, @cascalheira2010, @almeida2000].

The complete Data Dictionary with all variables, attributes, values and description, with reference to the consulted literature is available in our research compendium at https://doi.org/10.17605/OSF.IO/456EG, and can be also consulted in the Supplementary Materials (Table S1).

After data collection, the databases were imported into R environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. We include the entire R code used for all the analysis and visualizations contained in this paper in our online research compendium. To produce those files, we followed the procedures described by Marwick et al. [-@marwick2017] for the creation of research compendiums to enhance the reproducibility of research. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY (for more information, see @marwick2018]. All attribute analysis result tables may also be found in the Supplementary Materials.

@@

Vale Boi is an open-air site and rockshelter, located on the western coast of Algarve (southern Portugal), near a small homonymous village, within the municipality of Vila do Bispo (Figure \@ref(fig:protomap)). The site is situated in a small valley that runs south to the Atlantic coast, about 2 km distance, relatively open, bordered, to the east, by a limestone hill. This hill is marked, at specific points, by limestone exposures that form rock shelters with faces facing west or southwest [@bicho2003; @cascalheira2010; @cascalheiraetal2008].

The site extends for more than 10 000 $m^2$ on the slope of this valley, which is marked by a series of steps that run parallel to the river, possibly the result of Middle Pleistocene fluviatile erosion [@bicho2003].

The geological context of Vale Boi is marked by heterogeneity. In the north, there are schist and greywacke formations from the Carboniferous. In the south, Triassic and Jurassic dolomite and limestone formations, which are gradually covered by Holocene dunes further into the coastal area, until near St. Vincent’s cape where they appear uncovered once again, along with small occurrences of chert [@verissimo2004].

Vale Boi shows a variety of human occupations, distributed across the three main areas: Slope, Shelter, and Terrace.

Although excavated since 2000, in 2012, a new 8 $m^2$ area was open in the Terrace (rows H and I), to understand the stratigraphic sequence in more detail and assess the existence of older cultural horizons, attributable to an early Upper Paleolithic occupation.

From 2012 to 2016, six litho-stratigraphic layers have been identified (Table S2). In some of these layers, lateral sediment variations were identified, which were coded in the field through the concatenation of a letter to the layer number (*e.g.* 4E). In many of these cases, the isolation of this variation did not show any patterns in terms of spatial concentration of materials, although in others, like the 4E facies, the subdivision of the layer correlated with the spatial distribution of Vale Comprido technology.

In the new area, layers 1 and 2 represent Holocene levels, with the first being disturbed by recent agricultural processes. Layer 2, with a thickness of 25-30 cm shows a Neolithic occupation.

@@

Layer 4 is very similar to layer 3. However, it is separated from it by a gravel horizon. Similarly to layer 4, these layers have been subdivided regarding different degrees of sediment compaction and/or concentration of organic materials, showing two differing cultural horizons: Solutrean and Proto-Solutrean (limited to layer 4E).

@@

Regarding layers 4E and 5, although they are indeed layers with different sedimentary characteristics, they show similar technological and archaeological patterns, which led to the conclusion they were the same cultural horizon.The most evident of all is the presence of Vale Comprido points and blanks. This context was dated using a charcoal sample to c. 24.7-25.3 ka cal BP, and a shell sample to c. 23.7-24.1 ka cal BP (Figure \@ref(fig:c14VB) and Table S3).

That both layers are representative of the same cultural horizon is also the is the vertical constraint in the distribution of dolerite pieces, mostly concentrated on Layer 4E and upper levels of Layer 5, coinciding also with the higher concentration of lithic materials (Figure \@ref(fig:spatialvbfig)).

The bottom spits of layer 5, on the other hand, show a relatively dense concentration of *Littorina littoreia*, an unprecedented shell species at the site, and often associated with colder waters, having a high freezing tolerance [@murphy1979; @clarke2000]. This context was dated through a shell sample to c. 26.2-25.7 ka cal BP and thus within the time interval estimated for the HE2.

@@

Taking in consideration the models available for the Gravettian-Solutrean transition in western Europe, particularly, the differences proposed for raw material exploitation (i.e. high vs. low use of quartz) our first approach was to check for the existence of raw materials differences across the layers just defined. For this we calculated the percentages of the three most important raw materials by cubic meter of excavated sediment (Figure \@ref(fig:rmdispersion)) and plotted them by elevation. There seems to be significant differences in raw material distribution over time. From around 24.1 m upwards, there is an important shift in quartz and chert frequencies, the latter increasing more than 10%, and quartz dropping from c. 50% to nearly 30%.

This shift seems to be associated with other abovementioned changes in the Terrace sequence, such as an increase in the amount of lithic materials in top of Layer 5 and Layer 4E, but also the appearance of Vale Comprido technology. Additionally, these two moments are stratigraphically correlated with the two different chronological horizons defined above: the first, dated to c. 26 ka cal BP, at c. 23.9 m in elevation, and associated with higher frequencies of quartz, and the second, dated to c. 24.7 ka cal BP, at around 24.1 m, associated with higher frequencies of chert and a reduction in quartz presence.

@@

Given the chronological data, density of artifacts, and raw material preference patterns, it was decided that for this study the materials would be subdivided in two analytical units, to better understand any possible technological differences: **Lower 5**, including all artifacts with elevation values under 24.1; **Upper 5/4E**, including all artifacts with elevation values of 24.1 and above. The value 24.1 is arbitrary and was chosen for the reasons aforementioned.

A total of 26,703 pieces were analysed for both groups, 11,094 from the Lower 5 group and 15,609 for the Upper 5/4E (Tables \@ref(tab:general1) and \@ref(tab:general2)). Most of these are chips (70.62% from Lower 5 and 65.68% from Upper 5/4E), followed by shatters, which make up c. 21% and c. 24% of the groups, respectively. These extremely high numbers for debris are mostly the result of quartz use, which can be explained not only by on-site knapping of this raw material but mostly by its breakage patterns (especially when coarser), which typically produces more waste than in other raw materials.

@@

For Lower 5, cores and debitage products represent a cumulative frequency of c. 8%. For Upper 5/4E, debitage products represent nearly 10% of the assemblage. Complete blanks are the most represented class for both groups, with 1,474 identified pieces, 4.59% for Lower 5 and 6.18% for Upper 5/4E, followed by blank fragments, with an absolute count of 248 for Lower 5 and 336 for Upper 5/4E. Cores are also relatively frequent within these assemblages, with an absolute count of 123 cores, 46 for Lower 5 and 77 for Upper 5/4E.

During the analysis, 167 complete retouched pieces were identified, 55 on Lower 5, representing 0.5% of the group, and 112 on Upper 5/4E, representing 0.72% of the assemblage. Retouched piece fragments appear in much smaller numbers (n=7 for Lower 5 and n=16 for Upper 5/4E).

One of the most relevant raw materials in both of Vale Boi's analytical units used for this study is quartz (encompassing rock crystal, fine, medium and coarse qualities), which represents c. 43% of the total assemblage for the Upper 5/4E group, and c. 51% for Lower 5. Chert, on the other hand, represents c. 46% of the Upper 5/4E phase and 38% of the Lower 5. Finally, greywacke, dolerite, chalcedony, and other raw materials that due to their low frequencies were collapsed in the category “Other”, together represent c. 11% of both assemblages (Figure \@ref(fig:rmvb)).

@@

The number of chips and shatter in quartz are extremely high, both in its representativity within the raw material, with c. 94% for both Lower 5 and Upper 5/4E, and when compared to the relative frequencies of other raw materials.

There is a relatively large number of quartz blanks (n=281 for Lower 5 and n=438 for Upper 5/4E) compared to the rest of the classes, as well as cores (n=18 for Lower 5 and n=32 for Upper 5/4E) and retouched pieces (n=15 for Lower 5 and n=34 for Upper 5/4E).

Chert shows high frequencies of blanks (15.8% for Lower 5 and 19.6% for Upper 5/4E) when compared to every other class, excluding chips. The latter represents more than 50% of the chert assemblages, with an absolute frequency of 649 chips for Lower 5 and 1,158 for Upper 5/4E.

Although chert shows smaller absolute numbers for blanks (n=171 for Lower 5 and n=407 for Upper 5/4E) and blank fragments (n=90 for Lower 5 and n=155 for Upper 5/4E) compared to quartz, there is a large quantity of cores representing 2.1% of total chert in Lower 5 and 2% in Upper 5/4E. However, there is a noticeable difference in the ratio between chert blanks and quartz blanks in Lower 5 and Upper 5/4E, with the latter showing a higher frequency of chert blanks compared to Upper 5/4E.

Regarding retouched pieces, these represent more than 3% of chert for both phases, a number far higher than those in other raw materials. This shows that chert, although less present than quartz in blank absolute count (although barely in Upper 5/4E), might have been preferentially used for formal tool production.

Dolerite seems to appear only in layers 4E and 5 of the Terrace and Proto-Solutrean levels of the Slope area. The most striking characteristic of this raw material in the assemblages under study is the low presence of chips or shatter (n=1 each), but high frequencies of blanks (50% in Lower 5 and 60.8% in Upper 5/4E) and blank fragments (16.6% in Lower 5 and 13% in Upper 5/4E). Unlike Lower 5, in Upper 5/4E dolerite is represented by five retouched tools, some of which are Vale Comprido points. This raw material seems to have been used preferentially at Vale Boi for the production of these types of points, a pattern already observed in previous works [@marreiros2009], where three out of five Vale Comprido points were identified as dolerite but originally thought to be jasper.

The almost complete absence of cores, with only one core fragment identified in the Lower 5 group, and low quantity of debitage waste may be explained by the importation of finished pieces or blanks to the site, suggesting that, most likely, no knapping activities occurred with this raw material at the site.

@@

### Cores

>>>>

---
title: "The Gravettian-Solutrean transition in westernmost Iberia: new data from the sites of Vale Boi and Lapa do Picareiro"
author:
  - Joana Belmiro: 
      email: jnetobelmiro@gmail.com
      institute: [ICArEHB]
      correspondence: true
  - João Cascalheira:
      email: jmcascalheira@ualg.pt
      institute: [ICArEHB]
      correspondence: false
  - Nuno Bicho:
      email:
      institute: [ICArEHB]
  - Jonathan Haws:
      email:
      institute: [UL, ICArEHB]
institute:
  - ICArEHB: ICArEHB, Universidade do Algarve
  - UL: University of Louisville
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: bibliography.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  This study presents the analysis of the lithic assemblages from Layers 5 and 4E (Terrace) of the site of Vale Boi (southern Portugal) and Levels U and T from Lapa do Picareiro (central Portugal). We aimed to understand the technological patterns and raw material exploitation during the Gravettian-Solutrean transition in westernmost Iberia and test the traditional models with assemblages from recently excavated sites, while expanding the geographic range. Results show the existence of two discrete phases in each site. The first, with high frequency of quartz use for bladelet production, seems to reflect the presence of a Terminal Gravettian horizon, as defined by Almeida (2000).The second, with some significant differences between sites, attests the presence of Vale Comprido technology and lower quartz frequencies at Vale Boi, representing a Proto-Solutrean occupation; and the presence of a blade component in Lapa do Picareiro that, together with the respective absolute chronology, may be attributed to a Proto-Solutrean or an Early Solutrean horizon. In general terms, this study allowed to confirm that the Terminal Gravettian and the Proto-Solutrean are discrete phases across the transition, in agreement with the Three-Phase model presented by Zilhão [-@zilhao1997]. It further consolidates the expansion of similar techno-cultural patterns to southern Portugal, which may be explained by the significant changes in the dynamic of social networks [@cascalheiraandbicho2013].

@@

The transition from the Gravettian to the Solutrean and its correlation with the environmental changes occurred during the Heinrich Event 2 (HE2) at the onset of the Last Glacial Maximum (LGM – ca. 26.5-19 ka cal BP), continues to be a critical topic to understand Upper Paleolithic human ecodynamics in westernmost Europe.

However, most of what is known for this timeframe is geographically constricted and/or lacking good chronological markers (Cascalheira and Bicho, 2013). In southern France and the western edge of Iberia, for example, this period is marked by the presence of the so-called Proto-Solutrean (Renard, 2011; Zilhão 1997). This has been described as a transitional technocomplex to the Solutrean technologies, through a process of local development, but synchronous across southwestern Europe (Zilhão et al., 1999). Although present in and southern Iberia (Alcaraz-Castaño et al., 2013; Aura et al., 2009), evidence of typical Proto-Solutrean industries is yet to be found in different regions of Iberia, such as the case of Cantabria, where the transition is poorly characterized (Cascalheira et al., in press).

In Portugal, the Proto-Solutrean was recognized as a result of the study of materials from old excavations by Manuel Heleno in the Rio Maior region (central Portugal) in the 1930s and 1940s, and from new archaeological work in the same region beginning in the late 1980s. From both phases, several open-air and cave sites are now known, and only one other site in southern Portugal (Vale Boi) has revealed typical Proto-Solutrean assemblages (Figure \@ref(fig:protomap)).

@@

There, the Proto-Solutrean has been integrated into the Gravettian-Solutrean transition as part of a sequence of techno-typological modifications that could have occurred across two or three different stages (Zilhão, 1997, 1994; Zilhão et al., 1999).

The Two-stage model starts with a Final Gravettian phase, characterized by moderate use of quartz (~15%), production of truncated backed bladelets, and proto-magdalenian retouched blades (Zilhão, 1997). The second phase is the Proto-Solutrean, frequently characterized by a high percentage of quartz use (~30%), rare presence of backed bladelets, and the production of Vale Comprido points, the technocomplex’s index fossil (Almeida, 2000; Zilhão, 1997; Zilhão and Aubry, 1995). For the production of these tools, the Proto-Solutrean shows two different operative sequences: one for the production of Vale Comprido points, through the removal of elongated blanks with convergent profiles and thick platforms; another for the production of blades, of Gravettian tradition; and another for the production of bladelets, probably obtained through the exploitation of thick endscrapers or carinated elements (Zilhão, 1997). Vale Comprido points are described as robust pieces, often characterized by convergent shapes, triangular cross-sections, and plain platforms, often having a high elongation ratio, although not necessarily falling into the blade category (Zilhão and Aubry, 1995). These points are seen as an element of discontinuity with the previous technocomplex, where the organic points armed with microliths from the Gravettian are replaced by lithic points, with enough similarities to the pointes à face plane of the Middle Solutrean to be understood as a technological development (Zilhão, 2013).

The Three-stage model, on the other hand, maintains the Final Gravettian phase and its defining characteristics but subdivides the following Proto-Solutrean stage in two. As such, there is an intermediate stage characterized by the intensive use of quartz (~30%), which corresponds to Laugerie-Haute’s Aurignacian V, and named Terminal Gravettian. The third stage is the Proto-Solutrean, where quartz use diminishes. In this model, the Vale Comprido points and associated reduction sequence may appear in either of the last two stages (Zilhão, 1997). Alternatively, in the Two-phase model, the Terminal Gravettian is understood as a functional facies of the Proto-Solutrean phase, related to specialized occupations and production activities within the technocomplex.

Regardless of which model is accepted, a high degree of lithic technological variability seems to be present across the transition (Almeida, 2000; Zilhão, 1997; Zilhão et al., 1999). This variability has been interpreted by Cascalheira and Bicho (2013) as a technological race in response to the environmental modifications taking course during the HE2. In this framework, the authors suggest that, in order to correspond to the external pressures, there may have been the need to diversify the economic strategies in use until that moment. The high exploitation of quartz, for example, formerly a secondary raw material, and the use of similar reduction strategies between quartz and chert, might represent one such economic response. The same possibly applies to the use of new raw materials, as is the case of the Proto-Solutrean levels in Vale Boi, where dolerite starts to be used mainly for the manufacture of Vale Comprido points (see below and Belmiro et al., 2017; Marreiros, 2009).

However, despite the existence of a rather comprehensive set of literature about the technocomplex in central Portugal, which has allowed for a better understanding of  the transition between the Gravettian and the Solutrean, further studies are needed to fill the existing gaps. One of these gaps is the concentration of data regarding this technocomplex in the Portuguese Estremadura, whereas the Proto-Solutrean is still relatively unknown in other areas throughout southwestern Europe. In the case of Portugal, only one other site in the south (Figure 1) has so far revealed the presence of this techno-complex. Another issue is the small number of absolute dates available for the transition, which limits the chronological definition of the technocomplexes, and thus the testing of which of the abovementioned models (if any) applies best (Cascalheira and Bicho, 2013).

Addressing these issues, through the study of lithic assemblages from sites with good stratigraphic preservation, allowing accurate absolute dating and spatial tracking, will undoubtedly help further understand the Gravettian-Solutrean transition, its patterns, stages, and possible regional variations.

@@

By combining techno-typological information with stratigraphic data and absolute chronology, we present new data on: 1) the technological organization of levels 5 and 4E of Vale Boi, and levels U and T from Lapa do Picareiro; and 2) a stratigraphical and chronological separation between Terminal Gravettian horizons and more recent occupations at both sites, revealing the application of the same general set of patterns to geographic areas outside of the Portuguese Estremadura.

@@

For the lithic analysis we adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analyzed followed those present in specialized literature, such as Brézillon (1968), and Tixier and Inizan (1980), paired with other Paleolithic lithic attribute analyses (see e.g., Bicho, 1992; Zilhão, 1997; Cascalheira, 2019; Almeida, 2000; Scerri et al., 2014; Tostevin, 2013).

A complete data dictionary with all variables, attributes, and descriptions, with reference to the consulted literature, is available in our research compendium at https://www.doi.org/10.17605/OSF.IO/456EG, and can be also consulted in the Supplementary Materials (Table S1).

After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. We include the entire R code used for all the analyses and visualizations contained in this paper in our online research compendium. To produce those files, we followed the procedures described by Marwick et al. (2017) for the creation of research compendiums to enhance the reproducibility of research. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY (for more information, see Marwick et al. 2018).

@@

Vale Boi is an open-air site and rockshelter, located on the western coast of Algarve (southern Portugal), near a small homonymous village, within the municipality of Vila do Bispo (Figure \@ref(fig:protomap)). The site is situated in a small valley that runs south to the Atlantic coast, about 2 km distance, relatively open, bordered, to the east, by a limestone hill. This hill is marked, at specific points, by limestone exposures that form rock shelters with faces facing west or southwest (Bicho et al., 2003; Cascalheira, 2010; Cascalheira et al., 2008).

Human occupation extends for more than 10 000 sq. meters on the slope of this valley, which is marked by a series of steps that run parallel to the river, possibly as a result of Middle Pleistocene fluviatile erosion (Bicho et al., 2003).

The geological context of Vale Boi is marked by heterogeneity. In the north, there are schist and greywacke formations from the Carboniferous. In the south, Triassic and Jurassic dolomite and limestone formations, which are gradually covered by Holocene dunes further into the coastal area, until near St. Vincent’s cape where they appear uncovered once again, along with small occurrences of chert (Veríssimo, 2004).

Vale Boi shows a variety of human occupations, distributed across the three main excavation areas: Slope, Shelter, and Terrace (Bicho et al., 2012).

Although excavated since 2000, in 2012 a new 8 sq. meters area was open in the Terrace (rows H and I), to understand the stratigraphic sequence in more detail and assess the existence of older cultural horizons, attributable to an early Upper Paleolithic occupation.

From 2012 to 2016, six litho-stratigraphic layers have been identified (Table S2). In some of these layers, lateral sediment variations were detected, which were coded in the field through the concatenation of a letter to the layer number (e.g., 4E). In many of these cases, the isolation of this variation did not show any patterns in terms of spatial concentration of materials, although in others, like the 4E facies, the subdivision of the layer correlated with the spatial distribution of Vale Comprido technology.

In the new area, layers 1 and 2 represent Holocene levels, with the first being disturbed by recent agricultural processes. Layer 2, with a thickness of 25-30 cm, shows a Neolithic occupation.

@@

Layer 4 is very similar to layer 3. However, it is separated from it by a gravel horizon. Similarly, layer 4 have been subdivided regarding different degrees of sediment compaction and/or concentration of organic materials, showing two differing cultural horizons: Solutrean and Proto-Solutrean (limited to layer 4E).

@@

Regarding layers 4E and 5, although they are indeed layers with different sedimentary characteristics, they partially show similar technological and archaeological patterns, which led to the conclusion they in part represent similar cultural horizons. The most evident of all is the presence of Vale Comprido points and blanks in Layer 4E e the top of Layer 5. This context was dated using a charcoal sample to c. 24.7-25.3 ka cal BP, and a shell sample to c. 23.7-24.1 ka cal BP (Figure 3 and Table S3).

That both layers are representative of the same cultural horizon is also seen in the vertical constraint in the distribution of dolerite pieces, mostly concentrated on Layer 4E and upper levels of Layer 5, coinciding also with an higher concentration of lithic materials (Figure \@ref(fig:spatialvbfig)).

The bottom spits of Layer 5, on the other hand, show a relatively dense concentration of **Littorina littorea**, an unprecedented shell species at the site, and often associated with colder waters, having a high freezing tolerance (Clarke et al., 2000; Murphy, 1979). This context was dated through a shell sample to c. 26.2-25.7 ka cal BP and thus within the time interval estimated for the HE2 in westernmost Iberia (Sánchez-Goñi et al. 2008).

@@

Taking into consideration the models available for the Gravettian-Solutrean transition in westernmost Europe, particularly the differences proposed for raw material exploitation (i.e., high vs. low use of quartz) our first approach was to check for the existence of raw material differences across the layers just defined. For this, we calculated the percentages of the three most important raw materials by cubic meter of excavated sediment (Figure 4) and plotted them by elevation. There seems to exist significant differences in raw material distribution over time. From around 24.1 m upwards, there is an important shift in quartz and chert frequencies, the latter increasing more than 10%, and quartz dropping from c. 50% to nearly 30%.

This shift seems to be associated with other abovementioned changes in the Terrace sequence, such as an increase in the amount of lithic materials in the top of Layer 5 and Layer 4E, but also the appearance of Vale Comprido technology. Additionally, these two moments are stratigraphically correlated with the two different chronological horizons defined above: the first, dated to c. 26 ka cal BP, at c. 23.9 m in elevation, and associated with higher frequencies of quartz, and the second, dated to c. 24.7 ka cal BP, after around 24.1 m, associated with higher frequencies of chert and a reduction in quartz presence.

@@

Given the chronological data, density of artifacts, and raw material preference patterns, it was decided that for this study the materials would be subdivided into two analytical units, to better understand any possible technological differences: Lower 5, including all artifacts with elevation values under 24.1; Upper 5/4E, including all artifacts with elevation values of 24.1 and above. The value 24.1 is arbitrary and was chosen for the aforementioned reasons.

A total of 26,703 pieces were analyzed for both groups, 11,094 from the Lower 5 group and 15,609 for the Upper 5/4E (Tables 1 and 2). For both groups, more than 70% of the artifacts are debitage waste, which is mostly the result of quartz use and can be explained by on-site knapping and breakage patterns.

@@

Debitage products represent nearly 10% of both assemblages. Complete blanks are the most represented class, with 4.59% for Lower 5 and 6.18% for Upper 5/4E. Cores are also relatively frequent within these assemblages, with an absolute count of 123 cores, 46 for Lower 5, and 77 for Upper 5/4E. A total of 167 complete retouched pieces were identified, 55 on Lower 5 and 112 on Upper 5/4E.

The most relevant raw materials in both Vale Boi’s analytical units for this study are quartz (c. 43% of the total assemblage for the Upper 5/4E group, and c. 51% for Lower 5) and chert (c. 46% of the Upper 5/4E phase and 38% of the Lower 5; Figure 5).

@@

The number of chips and shatter in quartz is extremely high, both in its representativity within the raw material, with c. 94% for both Lower 5 and Upper 5/4E, and when compared to the relative frequencies of other raw materials. These are followed by blanks, cores, and retouched pieces, although in smaller numbers. Chert shows high frequencies of blanks (15.8% for Lower 5 and 19.6% for Upper 5/4E) when compared to every other class, excluding chips. The latter represents more than 50% of the chert assemblages. Cores represent in both groups around 2% of the total assemblage.

Retouched pieces represent more than 3% of chert for both phases, a number far higher than those in other raw materials. This shows that chert was preferentially used for formal tool production.

Dolerite is only present in layers 4E and 5 of the Terrace (Figure S1), but also in other Proto-Solutrean levels of the Slope area. The most striking characteristic of this raw material in the assemblages under study is the low presence of chips and shatter (n=1 each), but high frequencies of blanks (Table 1). Unlike Lower 5, in Upper 5/4E, dolerite is represented by five retouched tools, some of which are Vale Comprido technology (Figure 6).

```{r vct, fig.cap="Vale Boi - Vale Comprido points in dolerite, chert and chalcedony.", out.width = '70%', fig.align='center', echo=FALSE}

knitr::include_graphics("figure/prancha_vct.jpg")

The almost complete absence of dolerite cores, with only one core fragment identified in the Lower 5 group, and low quantity of debitage waste may be explained by the importation of finished pieces or blanks to the site, suggesting that, most likely, no knapping activities occurred with this raw material at the site.

@@

Excluding informal cores, there is a clear dominance of single platform cores, for most raw materials and in both phases (Tables S4 and S5). Chert presents the highest variability of core types (Figure 7), with high frequencies for single platform (50% for Lower 5 and 33.3% in Upper 5/4E), but also unidirectional prismatic on Lower 5.

Most cores were used for the extraction of flakes, although blade and bladelet scars also show relatively high frequencies associated with the exploitation of unidirectional prismatic cores (Figure S2).
